Report on the Second Workshop on Sustainable Software for Science: Practice and Experiences (WSSSPE2)

### 2015-07-05 WSSSPE2_report/WSSSPE2_review.tex
Section 9.1.3: minor changes.
Section 9-11: Minor changes.

\begin{itemize}
\item Journal Policies
\begin{itemize}
	\item Accessibility of papers
	\begin{itemize}
		\item Open Access
		\item ``Freely'' available
	\end{itemize}
	\item Repositories
	\begin{itemize}
		\item Provides suggestions for recommended repositories
		\item Provides its own repository
	\end{itemize}
	\item Review
	\begin{itemize}
		\item Reviewing software is mandatory
		\begin{itemize}
			\item Must check that software runs
			\item Must check quality of code
			\item Must check performance of code if paper makes claims on relative performance
		\end{itemize}
	\end{itemize}	
	\item Supporting Data
	\begin{itemize}
		\item Must be publicly available
		\item Must be in an Open Access repository
		\item must have a DOI
	\end{itemize}
	\item Article Processing Charge
	\begin{itemize}
		\item APC charge is transparent
		\item APC waiver program
	\end{itemize}
\end{itemize}

\item Paper Policies
\begin{itemize}
	\item Required Sections
	\item Keywords
	\begin{itemize}
		\item Paper provides keywords to help describe software
	\end{itemize}
	\item Papers can be updated when new releases of software are released
	\begin{itemize}
		\item At no extra cost/at significantly reduced cost
	\end{itemize}
\end{itemize}

@@

\subsubsection{Second breakout discussion: What should journals provide reviewers of software papers}

This group discussed whether they could come up with a list of things that a
journal should provide its reviewers to make it easier to review software
submitted for publication as a ``Software Paper''.

@@

Journals should provide mechanisms to enable and track communication between
reviewers and authors. For anonymous peer review journals, maintain anonymity of
reviewers. If communication is necessary, that may mean that software is not
that well documented. Journals should also provide a set of simple metrics for
software evaluation that reviewers can use for ratings, similar to Consumer
Reports.

@@

Journals should support mechanisms to run software. Sometimes this may be very
hard to accomplish, despite best efforts. Journals could provide an Execution
environment for any Software submitted for review, perhaps via a Docker
container or a virtual machine. If not, instructions must be adequate to compile
and execute the software; and interpret the results (output files, formats,
etc.), and the full source code must be accessible to the reviewer. Mechanisms
to quantify what has changed compared to a previously published version would
assist version comparison. However, this cannot simply take lines of code into
account. For example, a speedup of an algorithm may not result in a huge code
difference, but may nonetheless provide a break-through. Journals should require
software submissions to also provide `test materials' -- sample data, parameters
to validate that code is working as intended. In certain cases, well-selected
benchmark datasets may be required to assess performance and accuracy.

@@

This group began by recognizing that they know what the code testing and code
review problems are, so the purpose of the group was really to think about
concrete things that people in the group would be willing to commit to, in order
to improve the sustainability of software.

@@

A key point of discussion was that there are two major challenges to software
testing in science. The first is convincing developers to incorporate testing.
This challenge is both social and technical. We need to communicate the value of
testing and also teach developers how to choose and use testing frameworks. The
second challenge---and the one that the group felt was more difficult---is
choosing appropriate tests for scientific software. This is the difference
between learning how to use Python \texttt{unittest} and knowing how to test
that your code correctly implements, for example, a Lattice-Boltzmann model in
computational fluid dynamics.

@@

\item Formal tutorials at conferences (for example, a test driven development
(TDD) at a Software Developers Best Practices conference): These often require
self-learning following the tutorial, and it can be hard to find a tutorial that
matches the programming language and/or the domain of the participant.

\item As a way to be confident in other software: When breaking out pieces of
software from a larger application, a developer wants to ensure the software
works as expected. These types of tests may be ephemeral, though, living long
enough to give the author of the test confidence in the code, but not handed on
and made available to next user. These experiences can lead to more systematic
testing.

\item Through experience with coding: As you write more software, your
confidence in code is reduced and the importance of testing becomes clearer.
Attendees joked that they started testing ``when their skepticism/guilt became
larger than their arrogance''.

@@

Simpler, standard ways of setting up appropriate testing infrastructure will be
important for adoption by scientists. Jenkins
\footnote{\url{https://jenkins-ci.org/}} was suggested as a common open-source
solution, but the initial configuration was considered to be challenging for
typical scientists.

Teach testing be beginning with a (smallish) piece of code that lacks
appropriate tests and develop an exercise that involves refactoring the code
into a more presentable form through creating/deriving unit tests. Students
would extract unit-tests through reverse-engineering and/or questioning a
partner who is a developer/expert.

This lead into a more general discussion about whether programming courses
should be part of the standard curriculum for science students and the
challenges of fitting additional material into university curricula. This is the
reasons that short workshops such as Software Carpentry and others exist.

@@

\subsubsection{Writing tests for scientific software}
The second major discussion thread in the group addressed the issues identified
by Software Carpentry - the difficulties in writing tests appropriate for
scientific software. Even among this group, while many What tests are
appropriate to ensure that an complex method is working correctly? When is the
result of a numerical computation 'close enough' to pass a test?

Several members of the group indicated that while they valued the concept of
software testing in theory, they were unaware how to test certain kinds of
software. Or, more importantly, they lacked relevant examples of software
testing that would be suitable for new members to their team. As an example, the
group asked if it could produce a 100-line example of how to test neutron
transport (or other specific scientific examples) targeted at a sophomore? This
would be a demonstration of ``How do I do this right?''

Another question was ``How do I test so that I know that the code is not the
source of my problems?'' A common challenge is that of stripping things down to
an appropriate level for tests. The group recognized that the difficulty is
different for so-called software infrastructure than for numerical/scientific
layers of software applications.

%The group referred to Lorena Barba's ``12 Steps to Navier
%Stokes~\cite{12steps},'' an online CFD lesson repository created as a
%set of iPython notebooks.  \clunenote{I'm not sure of the context
%  here, except possibly as an analog for something under discussion.}

%
%Distinction between 100 lines of input files vs. 100 lines of source code \katznote{to delete? - don't see how it fits}
%\clunenote{I no longer recall the context for this statement; does not seem important to me.}

There is some discussion of the issue in the computer science
literature~\cite{hook1,hook2} but less in domain-specific journals. Some lessons
on multiphysics software verification can be found in a recent
paper~\cite{dubey-FLASH}.

Two important pieces of the barrier are a) picking/implementing testing
frameworks (a technical barrier) and b) deciding what are the actual tests that
I need to write? A possible method to solve the latter is to ask students ``how
is what you are doing different from typing random keys on the keyboard?'' and
then, turn their answer into the concept of tests. That is to say, expand the
process of creating code from one that is solely about ``how?'' into one that
also include ``why?''

@@

\item The size of the simplest piece of relevant, interesting code (i.e. an
estimate of lines of code)

\item A point-form list of the test cases you would use

@@

The differences in workshop organization in WSSSPE2 from WSSSPE1
are in using an existing service (EasyChair) to handle submissions and reviews,
rather than an ad hoc process, and using an existing service (well-sorted) to
allow collaborative grouping of papers into themes by all authors, reviewers,
and the community, rather than this being done in an ad hoc manner by the
organizers alone.

The fact remains that contributors also want to get credit for their
participation in the process. And the workshop organizers still want to make
sure that the workshop content and their efforts are recorded. Ideally, there
would be a service that would be able to index the contributions to the
workshop, serving the authors, the organizers, and the larger community. But
since there still isn't such a service today, the workshop organizers are
writing this initial report and making use of arXiv as a partial solution to
provide a record of the workshop.

@@

In terms of building community activities, we wanted to focus primarily on
working groups, which we were able to do, as discussed above, but we
also wanted to make sure that attendees felt they had a chance to get their
ideas across to the whole group, which was the purpose of the lightning talks.
Overall, this seemed to be successful at the time, in terms of both the lightning
talks and the breakout groups, and the discussion of sustainability also led
to interesting and useful results, but the challenge that we have discovered
since WSSSPE2 is that it is very hard to continue the breakout groups'
activities.  The WSSSPE2 participants were willing to dedicate their time to
the groups while they were at the meeting, but since then, they have gone
back to their (paid) jobs.  We need to determine how to tie the WSSSPE
breakout activities to people's jobs, so that they feel that continuing them
is a higher priority that it is now, perhaps through funding the participants,
or through funding coordinators for each activity, or perhaps by getting
the workshop participants to agree to a specific schedule of activities during the
workshop.

>>>>

\begin{itemize}
\item Journal Policies
\begin{itemize}
	\item Accessibility of papers
	\begin{itemize}
		\item Open access
		\item ``Freely'' available
	\end{itemize}
	\item Repositories
	\begin{itemize}
		\item Provides suggestions for recommended repositories
		\item Provides its own repository
	\end{itemize}
	\item Review
	\begin{itemize}
		\item Reviewing software is mandatory
		\begin{itemize}
			\item Must check that software runs
			\item Must check quality of code
			\item Must check performance of code if paper makes claims on relative performance
		\end{itemize}
	\end{itemize}	
	\item Supporting data
	\begin{itemize}
		\item Must be publicly available
		\item Must be in an open access repository
		\item must have a DOI
	\end{itemize}
	\item Article processing charge (APC)
	\begin{itemize}
		\item APC charge is transparent
		\item APC waiver program
	\end{itemize}
\end{itemize}

\item Paper Policies
\begin{itemize}
	\item Required sections
	\item Keywords
	\begin{itemize}
		\item Paper provides keywords to help describe software
	\end{itemize}
	\item Papers can be updated when new releases of software are available
	\begin{itemize}
		\item At no extra cost/at significantly reduced cost
	\end{itemize}
\end{itemize}

@@

\subsubsection{Second breakout discussion: What should journals provide reviewers of software papers?}

This group discussed whether they could come up with a list of things that a
journal should provide its reviewers to make it easier to review software
submitted for publication as a ``Software Paper.''

@@

Journals should provide mechanisms to enable and track communication between
reviewers and authors. For anonymous peer review journals, anonymity of
reviewers should be maintained. If communication is necessary, that may mean that software is not
that well documented. Journals should also provide a set of simple metrics for
software evaluation that reviewers can use for ratings, similar to Consumer
Reports.

@@

Journals should support mechanisms to run software. Sometimes this may be very
hard to accomplish, despite best efforts. Journals could provide an execution
environment for any software submitted for review, perhaps via a Docker
container or a virtual machine. If not, instructions must be adequate to compile
and execute the software; to interpret the results (output files, formats,
etc.); and the full source code must be accessible to the reviewer. Mechanisms
to quantify what has changed compared to a previously published version would
assist version comparison. However, this cannot simply take lines of code into
account. For example, a speedup of an algorithm may not result in a huge code
difference, but may nonetheless provide a breakthrough. Journals should require
software submissions to also provide `test materials' -- sample data, parameters
to validate that code is working as intended. In certain cases, well-selected
benchmark datasets may be required to assess performance and accuracy.

@@

This group began by recognizing that they knew what the problems of code testing and code
review are, so the purpose of the group was really to think about
concrete things that people in the group would be willing to commit to, in order
to improve the sustainability of software.

@@

A key point of discussion was that there are two major challenges to software
testing in science. The first is convincing developers to incorporate testing.
This challenge is both social and technical. We need to communicate the value of
testing and also teach developers how to choose and use testing frameworks. The
second challenge---and the one that the group felt was more difficult---is
choosing appropriate tests for scientific software. This is the difference
between learning how to use, say, Python \texttt{unittest} and knowing how to test
that one's code correctly implements, for example, a Lattice-Boltzmann model in
computational fluid dynamics.

@@

\item Formal tutorials at conferences (for example, a test driven development
(TDD) at a Software Developers Best Practices conference): These often require
self-learning following a tutorial, but it can be hard to find a tutorial that
matches the programming language and the domain of the participant.

\item As a way to be confident in other software: When breaking out pieces of
software from a larger application, a developer wants to ensure the software
works as expected. These types of tests may be ephemeral though, living long
enough to give the author of the test confidence in the code, but not handed on
and made available to next user. These experiences can lead to more systematic
testing.

\item Through experience with coding: As you write more software, your
confidence in code is reduced and the importance of testing becomes clearer.
Attendees joked that they started testing ``when their skepticism/guilt became
larger than their arrogance.''

@@

Simpler, standard ways of setting up appropriate testing infrastructure will be
important for adoption by scientists. Jenkins\footnote{\url{https://jenkins-ci.org/}} 
was suggested as a common open-source
solution, but the initial configuration was considered to be challenging for
typical scientists.

Teach testing by beginning with a (smallish) piece of code that lacks
appropriate tests and develop an exercise that involves refactoring the code
into a more presentable form through creating/deriving unit tests. Students
would extract unit-tests through reverse-engineering or questioning a
partner who is a developer/expert.

This leads to a more general discussion about whether programming courses
should be part of the standard curriculum for science students and the
challenges of fitting additional material into university curricula. This is the
reasons that short workshops such as Software Carpentry and others exist.

@@

\subsubsection{Writing tests for scientific software}
The second major discussion thread in the group addressed the issues identified
by Software Carpentry --- the difficulties in writing tests appropriate for
scientific software: %Even among this group, while many 
What tests are
appropriate to ensure that an complex method is working correctly? When is the
result of a numerical computation `close enough' to pass a test?

Several members of the group indicated that while they valued the concept of
software testing in theory, they were unaware of how to test certain kinds of
software. Or, more importantly, they lacked relevant examples of software
testing that would be suitable for members new to their teams. As an example, the
group asked if it could produce a 100-line example of how to test neutron
transport (or other specific scientific examples) targeted at a sophomore? This
would be a demonstration of ``How do I do this right?''

Another question was ``How do I test so that I know that the code is not the
source of my problems?'' A common challenge is that of stripping things down to
an appropriate level for tests. The group recognized that the difficulty is
different for so-called software infrastructure than for numerical/scientific
layers of software applications.
%
%The group referred to Lorena Barba's ``12 Steps to Navier
%Stokes~\cite{12steps},'' an online CFD lesson repository created as a
%set of iPython notebooks.  \clunenote{I'm not sure of the context
%  here, except possibly as an analog for something under discussion.}
%
%
%Distinction between 100 lines of input files vs. 100 lines of source code \katznote{to delete? - don't see how it fits}
%\clunenote{I no longer recall the context for this statement; does not seem important to me.}
%
There is some discussion of the issue in the computer science
literature~\cite{hook1,hook2} but less in domain-specific journals. Some lessons
on multiphysics software verification can be found in a recent
paper~\cite{dubey-FLASH}.

Two important pieces of the barrier are a) picking/implementing testing
frameworks (a technical barrier) and b) deciding what are the actual tests that
I need to write? A possible method to solve the latter is to ask students ``how
is what you are doing different from typing random keys on the keyboard?'' and
then, turn their answer into the concept of tests. That is to say, expand the
process of creating code from one that is solely about ``how?'' into one that
also includes ``why?''

@@

\item The size of the simplest piece of relevant, interesting code (i.e., an
estimate of lines of code)

\item A point-form list of the test cases you would use.

@@

The differences in workshop organization in WSSSPE2 from WSSSPE1
are in using an existing service (EasyChair) to handle submissions and reviews,
rather than an ad hoc process, and using an existing service (Well Sorted) to
allow collaborative grouping of papers into themes by all authors, reviewers,
and the community, rather than this being done in an ad hoc manner by the
organizers alone.

The fact remains that contributors want to get credit for their
participation in the process. The workshop organizers also want to make
sure that the workshop content and their efforts are recorded. Ideally, there
would be a service that could index the contributions to the
workshop, serving the authors, the organizers, and the larger community. 
Since there still isn't such a service today, the workshop organizers are
writing this initial report and making use of arXiv as a partial solution to
provide a record of the workshop.

@@

In terms of building community activities, we wanted to focus primarily on
working groups, which we were able to do, as discussed above, but we
also wanted to make sure that attendees felt they had a chance to get their
ideas across to the whole group, which was the purpose of the lightning talks.
Overall, this seemed to be successful at the time, in terms of both the lightning
talks and the breakout groups, and the discussion of sustainability also led
to interesting and useful results. However, the challenge that we have discovered
since WSSSPE2 is that it is very hard to continue the breakout groups'
activities.  The WSSSPE2 participants were willing to dedicate their time to
the groups while they were at the meeting, but afterwards, they have gone
back to their (paid) jobs.  We need to determine how to tie the WSSSPE
breakout activities to people's jobs, so that they feel that continuing them
is a higher priority than it is now, perhaps through funding the participants,
or through funding coordinators for each activity, or perhaps by getting
the workshop participants to agree to a specific schedule of activities during the
workshop.
